First Class Pattern Matching in a Deeply Embedded Domain Specific Language\\ \large{Not For Distribution}

### 2020-07-20 paper/paper.tex
Expanding on Church encoding derivation idea in Church encoding section of paper

More generally, an algebraic datatype can be thought of as a sum of products
(that is, a coproduct of products).

\begin{todo}
  Maybe it could make sense to essentially derive the pattern matching representation
  from this Church encoding. If so, this could be described around here.
\end{todo}

>>>>

The type signature of \ttt{church\_coProductMatchEval} can be rewritten in the same form as a
typical deep embedding evaluation function by moving the second argument into the right-hand
side of the definition:

\begin{lstlisting}
church_coproductMatchEval
  :: ChurchCoproductMatch t -> t

church_coproductMatchEval
    (ChurchCoproductMatch f g)
  = \ c ->
      case c of
        Left x  -> f x
        Right y -> g y
\end{lstlisting}

An embedding for product matches can be similarly derived from the Church encoding for product
types:

\begin{lstlisting}
data ChurchProductMatch t where
  ChurchProductMatch
    :: (a -> b -> r) ->
       ChurchProductMatch ((a, b) -> r)

church_productMatchEval
  :: ChurchProductMatch t -> t

church_productMatchEval
    (ChurchProductMatch f)
  = \ p ->
      case p of
        (x, y) -> f x y
\end{lstlisting}

As we will have another notion of function types embedded in our language, we define two
types to distinguish between these. We define two of them so that we can also distinguish
between coproduct matches and product matches.

\begin{lstlisting}
newtype CoproductMatch a r
  = CoproductMatch (a -> r)

newtype ProductMatch a r
  = ProductMatch (a -> r)

data ChurchCoproductMatch t where
  ChurchCoproductMatch
    :: (a -> r) -> (b -> r) ->
       ChurchCoproductMatch
         (CoproductMatch (Either a b) r)

church_coproductMatchEval
  :: ChurchCoproductMatch t -> t

church_coproductMatchEval
    (ChurchCoproductMatch f g)
  = CoproductMatch (\ c ->
      case c of
        Left x  -> f x
        Right y -> g y)
\end{lstlisting}

The definition for product follows similarly, using \\ \ttt{ProductMatch}.

Here we make the observation that algebraic datatypes are structured as coproducts of products. With
this information in hand, we can combine \ttt{ChurchCoproductMatch} and \ttt{ChurchProductMatch}.

\begin{lstlisting}
newtype CoproductMatch a r
  = CoproductMatch
      { runCoproductMatch :: a -> r }

newtype ProductMatch a r
  = ProductMatch
      { runProductMatch :: a -> r }

data ChurchMatch t where
  ChurchCoproductMatch
    :: ChurchMatch (ProductMatch a r) ->
       ChurchMatch (CoproductMatch b r) ->
       ChurchMatch
         (CoproductMatch (Either a b) r)

  ChurchProductMatch
    :: ChurchMatch (a -> ProductMatch b r)
       -> ChurchMatch (ProductMatch (a, b) r)

  ChurchOneCoproductMatch
    :: ChurchMatch (ProductMatch a r) ->
       ChurchMatch (CoproductMatch a r)

  ChurchOneProductMatch
    :: (a -> r) ->
       ChurchMatch (ProductMatch a r)

  ChurchNullaryProductMatch
    :: r -> ChurchMatch (ProductMatch a r)

  ChurchLam
    :: (a -> b) -> ChurchMatch (a -> b)

church_matchEval :: ChurchMatch t -> t
church_matchEval
    (ChurchCoproductMatch f g) =
  CoproductMatch (\ c ->
    case c of
      Left x  ->
        runProductMatch
          (church_matchEval f) x
      Right y ->
        runCoproductMatch
          (church_matchEval g) y)

church_matchEval
    (ChurchProductMatch f) =
  ProductMatch (\ p ->
    case p of
      (x, y) ->
        runProductMatch
          (church_matchEval f x) y)

church_matchEval
    (ChurchOneCoproductMatch f) =
  CoproductMatch (\ x ->
    runProductMatch
      (church_matchEval f) x)

church_matchEval
    (ChurchOneProductMatch f) =
  ProductMatch (\ x -> f x)

church_matchEval
    (ChurchNullaryProductMatch x) =
  ProductMatch (\ _ -> x)

church_matchEval
    (ChurchLam f) = \ x -> f x
\end{lstlisting}

% \begin{lstlisting}
% data ChurchMatch t where
%   ChurchCoproductMatch
%     :: ChurchMatch ((a, b) -> r) -> ChurchMatch (Either b c -> r) ->
%        ChurchCoproductMatch (Either
% \end{lstlisting}

% \begin{todo}
%   Maybe it could make sense to essentially derive the pattern matching representation
%   from this Church encoding. If so, this could be described around here.
% \end{todo}
